The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-02-07 pkg/yuimadocs/inst/doc/JSS/article-new.Rnw
various update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@225 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Quasi Maximum Likelihood Estimation}
{\bf ADD SOMETHING ON ASYMPT DISTRIBUTION \& SAMPLING SCHEME} 
Consider the multidimensional diffusion process
\begin{equation}
\label{eq:sdemle}
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t
\end{equation}
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial value $X_0=x_0$. 
Moreover, $\theta_1\in\Theta_1\subset \mathbb{R}^p, p\geq1$,
$\theta_2\in\Theta_2\subset \mathbb{R}^q, q\geq1$,
$a:\mathbb{R}^d\times\Theta_2 \to \mathbb{R}^d$ and $b:\mathbb{R}^d\times\Theta_1 \to \mathbb{R}^d\times \mathbb{R}^r$.

@@

Adaptive Bayes estimation is developed in \pkg{yuima} via the method $\tt adaBayes$.
Consider again the model \eqref{eq:model1} with the same values for the parameters.
In order to perform Bayesian estimation, we need to prepare the prior densities for the parameters.
For simplicity we use uniform distributions in $[0,1]$
<<echo=TRUE>>=
prior <- list(theta2=list(measure.type="code",df="dunif(theta2,0,1)"), theta1=list(measure.type="code",df="dunif(theta1,0,1)"))
@
Then we call $\tt adaBayes$, on the same sample data we used for the \code{qmle} function, as follows
<<echo=TRUE, results=hide>>=
bayes1 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we can compare the adaptive Bayes estimates  with the QMLE estimates
<<echo=TRUE>>=
coef(bayes1)
coef(mle1)
@
%The argument \code{method="nomcmc"} in \code{adaBayes} performs numerical
%integration, otherwise MCMC method is used.
\subsubsection{The Effect of Small Sample Size in Drift Estimation}
It is known from the theory that the estimation of the drift in a diffusion process strongly depend on the length of the observation interval $[0,T]$.
In our example above, we took $T=n^\frac13$, with $n = \Sexpr{n}$, which is approximatively  \Sexpr{round(n^(1/3),2)}. Now we reduce the sample size to $n=500$ and the value of $T$ is then $T=\Sexpr{round(500^(1/3),2)}$.
We then apply both quasi-maximum likelihood and adaptive Bayes type estimators to these data
<<>>=
n <- 500
ysamp <- setSampling(Terminal = n^(1/3), n = n)
yuima <- setYuima(model = ymodel, sampling = ysamp)
set.seed(123)
yuima <- simulate(yuima, xinit = 1, true.parameter = list(theta1 = 0.2, theta2 = 0.3))
param.init <- list(theta2=0.5,theta1=0.5)
mle2 <- qmle(yuima, start =param.init , 
lower = list(theta1=0, theta2=0), 
upper = list(theta1=1, theta2=1))
bayes2 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we look at the estimates
<<>>=
coef(bayes2)
coef(mle2)
@
Compared to the results above, we see that the parameters in the diffusion coefficients are estimated with good quality while for the parameters in the drift the quality of estimation deteriorates. The adaptive Bayes estimator seems to perform a little better though.

@@

\subsection{LASSO Model Selection}
Let $X_t$ be a  diffusion process solution to
$$
\de X_t = a( X_t,\alpha) \de t + b(X_t,\beta)  \de W_t
$$
$$\alpha=(\alpha_1,...,\alpha_{p})'\in\Theta_p\subset \mathbb{R}^p, \quad p\geq1$$
$$\beta=(\beta_1,...,\beta_q)'\in\Theta_q\subset \mathbb{R}^q, \quad q\geq1$$
with $b:\Theta_p\times\mathbb{R}^d\to \mathbb{R}^d$, $\sigma:\Theta_q\times \mathbb{R}^d\to \mathbb{R}^d\times \mathbb{R}^m$ and $W_t,t\in [0,T]$, is a standard Brownian motion in $\mathbb{R}^m$. 
We assume that the functions $a$ and $b$ are known up to   $\alpha$ and $\beta$.
We denote by $\theta=(\alpha,\beta)\in\Theta_p\times \Theta_q=\Theta$ the parametric vector and with $\theta_0=(\alpha_0,\beta_0)$ its unknown true value. 
Let $\mathbb{H}_n({\bf X}_n,\theta) = \ell_n({\bf X}_n,\theta)$ from equation \eqref{qlik}.
The quasi-MLE $\hat{\theta}$ for this model is the solution of the following problem
$$
\hat{\theta}=(\hat\alpha,\hat\beta)'=\arg\min_\theta \mathbb{H}_n({\bf X}_n,\theta)
$$ 
The adaptive LASSO estimator is defined as the solution to the quadratic problem under $L_1$ constraints
$$
\check{\theta}=(\check\alpha,\check\beta)=\arg\min_\theta\mathcal{F}(\theta).
$$
with
$$
\mathcal{F}(\theta)=(\theta-\hat{\theta})^T\ddot{\mathbb{H}}_n({\bf X}_n, \hat\theta)(\theta-\hat{\theta})+\sum_{j=1}^p\lambda_{n,j}|\alpha_j| +\sum_{k=1}^q\gamma_{n,k}|\beta_k|
$$
For more details see \citet{DegIac10b}.
The tuning parameters should be chosen as in \citet{Zou06} in the following way
\begin{equation}
\label{eq:penalty}
\lambda_{n,j} = \lambda_0 |\hat \alpha_{j}|^{-\delta_1}, \qquad
\gamma_{n,k} = \gamma_0 |\hat \beta_{j}|^{-\delta_2}
\end{equation}
where $\hat \alpha_{j}$ and  $\hat \beta_{k}$ are the unpenalized QML estimator of $\alpha_j$ and $\beta_k$ respectively, $\delta_1, \delta_2>0$ and usually taken unitary.

>>>>

\subsection{Quasi Maximum Likelihood Estimation}
{\bf NAKAHIRO: ADD SOMETHING ON ASYMPT DISTRIBUTION \& SAMPLING SCHEME} \par
Consider the multidimensional diffusion process
\begin{equation}
\label{eq:sdemle}
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t
\end{equation}
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial value $X_0=x_0$. 
Moreover, $\theta_1\in\Theta_1\subset \mathbb{R}^p, p\geq1$,
$\theta_2\in\Theta_2\subset \mathbb{R}^q, q\geq1$,
$a:\mathbb{R}^d\times\Theta_2 \to \mathbb{R}^d$ and $b:\mathbb{R}^d\times\Theta_1 \to \mathbb{R}^d\times \mathbb{R}^r$.

@@

Adaptive Bayes estimation is developed in \pkg{yuima} via the method $\tt adaBayes$.
Consider again the model \eqref{eq:model1} with the same values for the parameters.
In order to perform Bayesian estimation, we need to prepare the prior densities for the parameters.
For simplicity we use uniform distributions in $[0,1]$
<<echo=TRUE>>=
prior <- list(theta2=list(measure.type="code",df="dunif(theta2,0,1)"), theta1=list(measure.type="code",df="dunif(theta1,0,1)"))
@
Then we call $\tt adaBayes$, on the same sample data we used for the \code{qmle} function, as follows
<<echo=TRUE, results=hide>>=
bayes1 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we can compare the adaptive Bayes estimates  with the QMLE estimates
<<echo=TRUE>>=
coef(bayes1)
coef(mle1)
@
%The argument \code{method="nomcmc"} in \code{adaBayes} performs numerical
%integration, otherwise MCMC method is used.
\subsubsection{The Effect of Small Sample Size in Drift Estimation}
It is known from the theory that the estimation of the drift in a diffusion process strongly depend on the length of the observation interval $[0,T]$.
In our example above, we took $T=n^\frac13$, with $n = \Sexpr{n}$, which is approximatively  \Sexpr{round(n^(1/3),2)}. Now we reduce the sample size to $n=500$ and then $T=\Sexpr{round(500^(1/3),2)}$.
We then apply both quasi-maximum likelihood and adaptive Bayes type estimators to these data
<<>>=
n <- 500
ysamp <- setSampling(Terminal = n^(1/3), n = n)
yuima <- setYuima(model = ymodel, sampling = ysamp)
set.seed(123)
yuima <- simulate(yuima, xinit = 1, true.parameter = list(theta1 = 0.2, theta2 = 0.3))
param.init <- list(theta2=0.5,theta1=0.5)
mle2 <- qmle(yuima, start =param.init , 
lower = list(theta1=0, theta2=0), 
upper = list(theta1=1, theta2=1))
bayes2 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we look at the estimates
<<>>=
coef(bayes2)
coef(mle2)
@
Compared to the results above, we see that the parameters in the diffusion coefficients are estimated with good quality while for the parameters in the drift the quality of estimation deteriorates. The adaptive Bayes estimator seems to perform a little better though.

@@

\subsection{LASSO Model Selection}
{\bf: STEFANO: EXPLAIN WHAT LASSO MEANS}\par
Let $X_t$ be a  diffusion process solution to
$$
\de X_t = a( X_t,\alpha) \de t + b(X_t,\beta)  \de W_t
$$
$$\alpha=(\alpha_1,...,\alpha_{p})'\in\Theta_p\subset \mathbb{R}^p, \quad p\geq1$$
$$\beta=(\beta_1,...,\beta_q)'\in\Theta_q\subset \mathbb{R}^q, \quad q\geq1$$
with $b:\Theta_p\times\mathbb{R}^d\to \mathbb{R}^d$, $\sigma:\Theta_q\times \mathbb{R}^d\to \mathbb{R}^d\times \mathbb{R}^m$ and $W_t,t\in [0,T]$, is a standard Brownian motion in $\mathbb{R}^m$. 
We assume that the functions $a$ and $b$ are known up to   $\alpha$ and $\beta$.
We denote by $\theta=(\alpha,\beta)\in\Theta_p\times \Theta_q=\Theta$ the parametric vector and with $\theta_0=(\alpha_0,\beta_0)$ its unknown true value. 
Let $\mathbb{H}_n({\bf X}_n,\theta) = \ell_n({\bf X}_n,\theta)$ from equation \eqref{qlik}.
The quasi-MLE $\hat{\theta}$ for this model is the solution of the following problem
$$
\hat{\theta}=(\hat\alpha,\hat\beta)'=\arg\min_\theta \mathbb{H}_n({\bf X}_n,\theta)
$$ 
The adaptive LASSO estimator is defined as the solution to the quadratic problem under $L_1$ constraints
$$
\check{\theta}=(\check\alpha,\check\beta)=\arg\min_\theta\mathcal{F}(\theta).
$$
with
$$
\mathcal{F}(\theta)=(\theta-\hat{\theta})^T\ddot{\mathbb{H}}_n({\bf X}_n, \hat\theta)(\theta-\hat{\theta})+\sum_{j=1}^p\lambda_{n,j}|\alpha_j| +\sum_{k=1}^q\gamma_{n,k}|\beta_k|
$$
For more details see \citet{DegIac10b}.
The tuning parameters should be chosen as in \citet{Zou06} in the following way
\begin{equation}
\label{eq:penalty}
\lambda_{n,j} = \lambda_0 |\hat \alpha_{j}|^{-\delta_1}, \qquad
\gamma_{n,k} = \gamma_0 |\hat \beta_{j}|^{-\delta_2}
\end{equation}
where $\hat \alpha_{j}$ and  $\hat \beta_{k}$ are the unpenalized QML estimator of $\alpha_j$ and $\beta_k$ respectively, $\delta_1, \delta_2>0$ and usually taken unitary.
